---
title: "Tools for Data Manipulation"
author: 
  name: Sebastian Warnholz
  company: INWT-Statistics
output: 
  ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Motivation

  - `dplyr` and `data.table`
      - performance
      - idiomatic syntax
  - syntax may help, but
      - not with package development
      - you rewrite everything after running `R CMD check`
  - integration with custom class definitions
      - is not supported
  - syntactic sugar for anonymous functions (see for example `rlist` and
  recently `purrr`)


## The 'Solution'

  - extensive use of the formula class
  - pro: 
      - `R CMD check` friendly
      - not relying on non-standard-evaluation: normal dispatch and argument
      passing possible
  - contra:
      - using formulas is a matter of preference
      - unexplored so far


## Aim of the Package

  - don't loose the performance improvements from `dplyr` or `data.table`
  - reuse as much as possible
  - be generic
  - respect custom classes
  - abstract (my) typical (data manipulation) patterns


## Design Issues

  - `dplyr`: functions do one thing only!
      - `slice` and `filter`
      - manual dispatch on the second argument
  - `data.table`: let's write one function only!
      - `[.data.table` has about 1500 lines of code
      - you use `dat[]` as the interface to all features 
  - `dat`: relies on method dispatch
      - is somewhere in between
      - preserves the class (S3 + S4) but uses (mostly) `dplyr` in the background
      - `dplyr` now includes a link to `data.table`


## Examples: rows

```{r echo=TRUE}
data("airquality")
suppressPackageStartupMessages(library("dat"))
airQuality <- as.DataFrame(airquality)
airQuality[~Day < 7, ]
```


## Examples: rows

```{r echo=TRUE}
suppressPackageStartupMessages(library("data.table"))
as.data.table(airquality) %>% mutar(~ Day < 7)
```


## Examples: cols

```{r eval = FALSE, echo = TRUE}
airQuality[is.numeric]
# airquality[sapply(airquality, is.numeric)]

as.data.table(airquality) %>% mutar("Ozone:Wind") # remeber dplyr::select

mutar(airquality, "^Wi") # regex is triggered by '^'
airQuality["^Wi"]
# airquality[grepl("^Wi", names(airquality))]

airQuality[1] # error - no positional matching!

# Yes, base R syntax is otherwise okay:
ind <- 1:5
airQuality[ind, "Wind"]
```


## Examples: mutate + summary

```{r echo = TRUE, eval = FALSE}
airquality %>%
  mutar(
    meanWind ~ mean(Wind), 
    by = "Month"
  )

airQuality[meanWind ~ mean(Wind), by = "Month"]
```

- `mutar` can do
    - row subsetting
    - col subsetting (character + logical, regex, functions)
    - create new columns using `dplyr::mutate`
    - data aggregation using `dplyr::summarise` (triggered with `by`)
    - stops here


## Integration of Custom Classes

```{r error=TRUE, echo=TRUE}
setClass("MyData", contains = "data.table")
MyData <- function(...) new("MyData", data.table(...))
dat <- do.call(MyData, airquality)
dat[1, ] # it is just broken - same for dplyr
```


## `mutar` can do more!

```{r echo=TRUE}
setMethod("[", "MyData", mutar)
dat[1, ]
dat[1, CopyOfDay ~ Day]
```


## Additional Functions

- `map`: as a replacement for `lapply` and `mapply`
- `flatmap`: instead of `sapply`
- `replace`: to replace values in a vector
- `extract`: to extract values from a vector


- these functions introduce syntactic sugar for anonymous functions 
- they are generic functions and can be extended for new data types
- they have special awareness of `data.frame`s and `list`s


## Be `data.frame` aware

```{r echo = TRUE}
ind <- sapply(airquality, is.integer)
airquality[ind] <- lapply(
  airquality[ind], 
  function(x) as.character(x + 1)
)

map(airquality, x ~ as.integer(x) - 1L, is.character) %>%
  str
```


## Be `list` aware

```{r}
l <- list(
  a = list(name = "Me", age = 29),
  b = list(name = "You", age = ">30")
)
```

```{r echo=TRUE, eval=FALSE}
l <- list(
  a = list(name = "Me", age = 29),
  b = list(name = "You", age = ">30")
)

lapply(l, function(x) x["name"])
map(l, "name")
flatmap(l, "name")
```

```{r echo=TRUE, eval = TRUE}
flatmap(l, ~ .$name) # function(.) .$name
```


## Be paranoid  

- add runtime type checks

```{r echo=TRUE, error=TRUE}
flatmap(l, numeric : x ~ x$age)
flatmap(l, character(1) : x ~ x$name)
```


## Be comprehensive

```{r echo = TRUE}
extract(1:20, ~ . %% 2 == 0)
replace(1:10, ~ . %% 2 == 0, NA_real_)

vec <- 1:20
vec[vec %% 2 == 0]
```


## Be comprehensive

```{r eval = FALSE, echo = TRUE}
# non pipepable state of the art:
l <- list(x = 1, x1 = 2)
l$x <- NULL
l[grepl("^x", names(l))] <- NULL
l[sapply(l, is.numeric)] <- NULL
l[sapply(l, function(el) any(el == 1))] <- NULL

# suggestion:
replace(l, "x", NULL)
replace(l, "^x", NULL)
replace(l, is.numeric, NULL)
replace(l, el ~ any(el == 1), NULL)
```

- here I need tp parse in my head:
    - `$`, `[`, `<-`, `names`, `sapply`, `grepl`
    - vs. `replace`


## Summary

- you can fall back to or integrate `dplyr` and `data.table` functions
- syntactic sugar for anonymous functions
- use custom data types:
    - everything is a method: `map`, `replace`, and `extract`
    - `mutar` respects the data it operates on (for better or worse)
- have *pipeable* syntax: no replacement functions
- may not replace everything in an interactive data analysis `->` it covers
almost all (my) use cases for production code
- general *split-apply-combine* idiom is implemented as flatmap: I only talked
about the good parts

