---
title: "Tools for Data Manipulation"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Motivation

  - `dplyr` and `data.table`
  - syntax is a plus
      - rewrite everything after running `R CMD check`
  - performance
  - integration with custom class definitions
      - not supported
  - improve and integrate the definition of anonymous functions (see for example `rlist` and recently `purrr`)


## The 'Solution'

  - extensive use of the formula class
  - pro: 
      - `R CMD check` friendly
      - not relying on non-standard-evaluation: normal dispatch and argument passing possible
  - contra:
      - using formulas is a matter of preference


## Aim of the Package

  - don't loose the performance improvements from `dplyr` or `data.table`
  - reuse as much as possible
  - be generic
  - respect custom classes
  - abstract (my) typical (data manipulation) patterns


## Design Issues

  - `dplyr`: functions do one thing only!
      - `slice` and `filter`
  - `data.table`: let's write one function only!
      - `[.data.table` has about 1500 lines of code
      - 400 open issues on github
  - `dat`: I don't believe in extremes
      - is somewhere in between and uses methods
      - preserves the class (S3 + S4) and uses (mostly) dplyr in the background


## Examples: rows

```{r echo=TRUE}
data("airquality")
suppressPackageStartupMessages(library("dat"))
airQuality <- as.DataFrame(airquality)
airQuality[~Day < 7, ]
```


## Examples: rows

```{r echo=TRUE}
suppressPackageStartupMessages(library("data.table"))
as.data.table(airquality) %>% mutar(~Day < 7)
```


## Examples: cols

```{r eval = FALSE, echo = TRUE}
airQuality[is.numeric]
# airquality(sapply(airquality, is.numeric))

as.data.table(airquality) %>% mutar("Ozone:Wind") # remeber dplyr::select

mutar(airquality, "^Wi") # regex is triggered by '^'
airQuality["^Wi"]
# airquality[grepl("^Wi", names(airquality))]

airQuality[1] # error - no positional matching!

# Yes, base R syntax is otherwise okay:
ind <- 1:5
airQuality[ind, "Wind"]
```


## Examples: mutate + summary

```{r echo = TRUE, eval = FALSE}
airquality %>%
  mutar(
    meanWind ~ mean(Wind), 
    by = "Month"
  )

airQuality[meanWind ~ mean(Wind), by = "Month"]
```

- `mutar` can do
    - row subsetting
    - col subsetting (charater + logicals, regex, functions)
    - create new columns using `dplyr::mutate`
    - data aggregation using `dplyr::summarise` (triggered with `by`)
    - stops here


## Integration of Custom Classes

```{r error=TRUE, echo=TRUE}
setClass("MyData", contains = "data.table")
MyData <- function(...) new("MyData", data.table(...))
dat <- do.call(MyData, airquality)
dat[1, ] # it is just broken - same for dplyr
```


## `mutar` can do more!

- It is supposed to be used for custom class definitions:

```{r echo=TRUE}
invisible(setMethod("[", "MyData", mutar))
dat[1, ]
dat[1, CopyOfDay ~ Day]
```


## Additional functions

- `map`: as a replcament for `lapply` and `mapply`
- `flatmap`: instead of `sapply`
- `replace`: to replace values in a vector
- `extract`: to extract values from a vector

## Be `data.frame` aware

```{r echo = TRUE}
ind <- sapply(airquality, is.integer)
airquality[ind] <- lapply(airquality[ind], function(x) as.character(x + 1))
map(airquality, f(x) ~ as.integer(x) - 1L, is.character) %>%
  str
```


## Be `list` aware

```{r}
l <- list(
  a = list(name = "Me", age = 29),
  b = list(name = "You", age = ">30")
)
```

```{r echo=TRUE, eval=FALSE}
l <- list(
  a = list(name = "Me", age = 29),
  b = list(name = "You", age = ">30")
)

lapply(l, function(x) x["name"])
```

```{r echo=TRUE, eval = FALSE}
map(l, "name") # use regex or vectors
```

```{r echo=TRUE, eval = FALSE}
flatmap(l, "name")
```


## Be paranoid  
